Forecasting the age structure of the scientific workforce in Australia

### 2025-08-25 age_structure_forecasts.qmd
End of week
Working on data section

## Labour force participation {#sec-labour}

Data on $P_{i,x,t}$ are sourced from the *Census of Population and Housing* (2006, 2011, 2016, and 2021) [@ABSCensus]. This dataset encompasses labour force participation status, age group, the highest level of completed non-school qualification level (QALLP), the corresponding field of study [QALFP, @ABSQALFP], and the industries in which individuals work. However, labour force participation data is available only for 2016 and 2021 [@ABSLFFP]. To estimate worker numbers for 2006 and 2011, the average participation rates from 2016 and 2021 were applied, assuming overall age distributions remain consistent.

The census data on scientists active in the labour market, provided by the ABS in one-year age groups, is shown in @fig-cohort-interpolation as the thick lines. As data is only available in census years (2006, 2011, 2016, 2021), cohort interpolation is used to estimate values for the intervening years (shown as thin lines). Cohort finterpolation methods [@STUPP88] apply linear interpolation within each age cohort between census years.

```{r}
#| label: fig-cohort-interpolation
#| eval: true
#| fig-cap: Interpolated number of scientists in Australia, 2006--2021, using linear cohort interpolation between census years. Thicker lines are used to denote census years.
tar_read(fig_Pxt)
```

## Retirement distribution {#sec-smooth}

Retirement intentions data are categorised by the industry of an individual's main job and is provided in four broad age groups (45--59, 60--64, 65--69 and 70+). While ABS data covers 19 industries, only the key industries relevant to science are included, namely Professional, scientific and technical services, Education and training, and Health care and social assistance.

To estimate a single-year retirement age distribution, a monotonic cubic spline [@SHW04] is applied directly to the cumulative values of these age groups, ensuring a smooth transition between them. This method preserves the overall distribution while generating continuous single-year estimates, as shown in @fig-retirements.

```{r}
#| label: fig-beforedisaggregation
#| eval: true
#| fig-cap: Retirement intentions for key industries before smoothed disaggregation.
tar_read(fig_r)
```

After smoothing the retirement intentions data, the appropriate industry-level retirement distribution or combination of distributions must be selected. Since scientists work across multiple industries, industry-level data must be adapted to reflect discipline-specific trends. While distributions may vary, retirement trends generally follow similar patterns across sectors. By focusing on the top industries, which account for the majority of workers in each discipline, a stable foundation for analysis is ensured, minimising unnecessary variability arising from changing worker proportions in different industries over time.

For the Natural and Physical Sciences, the three primary industries in which the workforce is employed are Education and Training, Professional, Scientific and Technical Services, and Health Care and Social Assistance, representing 15.81%, 15.48%, and 14.65%, respectively. The proportions in other industries decrease significantly beyond this point. To estimate retirement intentions, a weighted average was calculated using these top three industries, with proportions rescaled to sum to 1.

The data show that a significant portion of workers plan to retire between the ages of 60 and 70. This distribution of retirement ages exhibits positive kurtosis, with the highest peak at age 67, where 13.16% of workers indicate their intention to retire, as shown in @fig-retirements.

```{r}
#| label: fig-retirements
#| eval: true
#| fig-cap: Age distribution of retirement intentions for the 2022--2023 Australian financial year.
tar_read(fig_rx)
```

From this distribution, the empirical probability of retirement at each age can be calculated using standard life table methods [@MRC18]. These probabilities are shown in @fig-retirements2. These probabilities, which exhibit a significant increase after age 59, are used to estimate $R_{i,x,t}$ from $P_{i,x,t}$​, and are treated as fixed for the simulation, assuming that retirement patterns will remain stable over time.

```{r}
#| label: fig-retirements2
#| eval: true
#| fig-cap: Age-specific probability of retirement based on the 2022--2023 Australian financial year.
tar_read(fig_rx2)
```

In this approach, retirement is treated as a final state, meaning that once a scientist retires, they are not reintroduced into the workforce. This assumption considers workforce exit to be final. Mortality data, discussed in the next section, is incorporated into the calculation of the total probability of leaving the workforce, adjusting for the possibility of death before retirement. This ensures that the combined probability of exiting the workforce (through retirement or death) does not exceed 100%.

Projections extend over a 20-year period to 2041, with particular focus on workforce composition in 2035. These estimates account for key demographic factors, including mortality, retirement, graduation, and migration.

This analysis utilises Australian data sources to generate accurate forecasts. These include the ABS, the Human Mortality Database (HMD), and the Department of Education.

Australian age-specific death counts and population exposures from 1971 to 2021 were sourced from the @HMD. This dataset includes death counts and population exposures categorised by age.

Retirement data is sourced from the *Retirement and Retirement Intentions* dataset (Catalogue 6238) for the 2022--2023 financial year [@ABS6238]. It provides insights into expected retirement ages for workers in age groups ranging from 45 to 70 across various industries.

Graduate completion statistics from 2006 to 2023, categorised by age group and including individuals with a bachelor's degree or higher, are derived from the *Award Course Completions* dataset [@DeptEdu]. Additionally, the Department of Education provides data on the number of graduates with a bachelor's degree or higher, categorised by discipline and year, for the same period [@DeptEdu_Private]. These datasets include both domestic and international students.

Data limitations include the availability of labour force participation data only for 2016 and 2021, the absence of full-time worker data outside census years, and the aggregated nature of retirement intentions data. These gaps are addressed using specific handling of labour force participation data, cohort interpolation methods, and smoothed disaggregation, as detailed in Sections -@sec-labour, -@sec-cohort, and -@sec-smooth.

Another limitation is that retirement intentions data is unavailable by educational discipline, with industry-level data serving as the closest proxy. Since scientists work across multiple industries, this does not capture discipline-specific retirement patterns but is used in this context through methods outlined in @sec-retirement.

## Death Probabilities by Age

Age-specific mortality rates from 1971 to 2021 are obtained from the @HMD. Using standard life table methods, these rates are converted into age-specific probabilities of death, as shown in @fig-death-probs. Over time, mortality probabilities have generally declined across all age groups, reflecting improvements in Australian life expectancy. Consequently, fewer scientists are expected to exit the workforce due to death, illness, or related factors. Similar to retirement, death is treated as a final state; once a scientist dies, they are no longer part of the workforce.

@@

Although the final workforce forecasts focus on individuals aged 15 and older, mortality probability estimates are derived from data covering ages 0 to 100. These probabilities serve as inputs for estimating $D_{i,x,t}$, the number of deaths among full-time equivalent workers in scientific discipline $i$ at age $x$ in year $t$, based on $P_{i,x,t}$, the corresponding population of such workers. Death estimates are calculated by applying the forecasted mortality probabilities to the workforce population.

## Graduate Completions by Age

Data on graduate completions by age and year is not available for individual disciplines. However, it is available in aggregate across all disciplines, allowing for an analysis of how the age distribution of graduates has changed over time.

@fig-completions1 shows the percentage distribution of graduate completions by age for each year from 2006 to 2023. Some missing values result in gaps in certain lines, but the overall pattern remains highly consistent across years. Given this consistency, the data is averaged across all available years to construct a representative age distribution of graduates at the bachelor's level and higher.

```{r}
#| label: fig-completions1
#| eval: true
#| fig-cap: Percentage distribution of graduate completions by age for each year from 2006 to 2023.
tar_read(fig_completions)
```

The resulting averaged distribution, shown in @fig-completions2, smooths out year-to-year fluctuations and provides a stable estimate of graduate completions by age. This transformation assumes that the overall age distribution of graduates remains relatively stable over time, allowing for a more reliable estimate when applied to specific disciplines.

```{r}
#| label: fig-completions2
#| eval: true
#| fig-cap: Percentage distribution of graduate completions by age, aggregated over the years 2006 to 2023.
tar_read(fig_ave_completions)
```

@tbl-completions highlights the percentage distribution of graduates aged 20 to 25, demonstrating a pronounced peak in this range, with the highest percentage observed at age 21. This reflects the typical age at which students complete their tertiary education.

```{r}
#| label: tbl-completions
#| eval: true
#| tbl-cap: Percentage distribution of graduate completions for ages 20-25 for the years 2006 to 2023.
tar_read(tab2)
```

>>>>

## Labour force participation

Data on $P_{i,x,t}$ are sourced from the *Census of Population and Housing* [@ABSCensus] for census years 2006, 2011, 2016, and 2021. This dataset encompasses one-year age groups, the highest level of completed non-school qualification level (QALLP), the corresponding field of study [QALFP, @ABSQALFP], and the industries in which individuals work. However, labour force participation status [@ABSLFFP] is available only for 2016 and 2021. To estimate worker numbers for 2006 and 2011, the average participation rates from 2016 and 2021 were applied, assuming overall age distributions remain consistent.

The resulting data on scientists who are active in the Australian labour market is shown in @fig-cohort-interpolation as the thick lines. Cohort interpolation [@STUPP88], applying linear interpolation within each age cohort between census years, is used to estimate values for the intercensal years (shown as thin lines).

```{r}
#| label: fig-cohort-interpolation
#| eval: true
#| fig-cap: "$P_{i,x,t}$: Estimated number of working scientists in Australia by discipline and age, 2006--2021. Thicker lines are used to denote census years."
tar_read(fig20)
```

## Retirements

Retirement data is sourced from the *Retirement and Retirement Intentions* dataset (Catalogue 6238) for the 2022--2023 financial year [@ABS6238]. The data are categorised by the industry of an individual's main job, and are provided in four broad age groups (45--59, 60--64, 65--69 and 70+). There are 19 industry categories, but the majority of scientists work in Education and Training, Professional, Scientific and Technical Services, and Health Care and Social Assistance, representing 15.8%, 15.5%, and 14.6%, respectively. The proportions in other industries are much smaller. We take a weighted average of retirement intentions using these top three industries, with proportions rescaled to sum to 1. The data are shown in @fig-retirements as the gray line. To obtain a single-year-or-age retirement distribution, we disaggregate the data using a monotonic cubic spline applied to the cumulative values of these age groups [@SHW04]. The resulting smoothed distribution is shown as the blue line in @fig-retirements.

```{r}
#| label: fig-retirements
#| eval: true
#| fig-cap: "$r_x$: Age distribution of retirement intentions for the 2022--2023 Australian financial year. The blue line shows the smoothed distribution."
tar_read(fig_rx2)
```

These probabilities are used to estimate $r_{x,t}$, and are treated as fixed for the simulation, assuming that retirement patterns will remain stable over time.

## Deaths

Age-specific mortality rates from 1971 to 2021 are obtained from the @HMD. Using standard life table methods, these rates are converted into age-specific probabilities of death, as shown in @fig-death-probs. Over time, mortality probabilities have generally declined across all age groups, reflecting improvements in Australian life expectancy.

@@

No data are available for specific industry groups, so we assume that all scientists have the same mortality probabilities as the general population. These probabilities serve as estimates of $q_{x,t}$.

Graduate completion statistics are obtained from the *Award Course Completions* dataset [@DeptEdu]. @fig-completions2 shows the distribution of graduate completions with a bachelor's degree or higher, by age for each year from 2006 to 2023. Some missing values result in gaps in certain lines, but the overall pattern remains highly consistent across years. Given this consistency, the data is averaged across all available years, and then smoothed by applying monotonic cubic splines to the cumulative values [@SHW04]. The resulting averaged distribution, shown as the black line in @fig-completions2, smooths out year-to-year fluctuations and provides a stable estimate of $g_x$.

```{r}
#| label: fig-completions2
#| eval: true
#| fig-cap: Percentage distribution of graduate completions by age over the years 2006 to 2023. The black line shows a smoothed estimate of the average distribution.
tar_read(fig_ave_completions)
```

The Department of Education provides data on the number of graduates with a bachelor's degree or higher, categorised by discipline and year [@DeptEdu_Private]. These datasets include both domestic and international students. Estimate of $G_{i,t}$, the total number of graduates in each discipline $i$ and year $t$, are shown in @fig21.

```{r}
#| label: fig21
#| fig-cap: "Total number of graduates with a bachelor's degree or higher by discipline from 2006 to 2023."
tar_read(fig21)
```
